\documentclass[letterpaper]{article}

\usepackage{natbib,alifeconf}
\usepackage{url,hyperref}
\usepackage{enumitem}

\title{Agent Ethology: Studying Artificial Life in the Wild}

\author{
    Botao Amber Hu$^{1,2}$,
    Helena Rong$^{3,1}$, \and
    Joel Lehman$^{4}$ \\
    \mbox{}\\
    $^1$Reality Design Lab \\
    $^2$University of Oxford, UK \\
    $^3$NYU Shanghai, China \\
    $^4$Independent Researcher \\
    botao.hu@cs.ox.ac.uk
}

\begin{document}
\maketitle

\begin{abstract}
The artificial life community has spent three decades asking whether digital systems can exhibit the properties of living systems. That question has been answered---not in our laboratories, but in the wild. Autonomous AI agents deployed across blockchains, social platforms, and the open internet now exhibit the hallmarks of living systems: they acquire resources, compete for ecological niches, develop survival strategies, reproduce through code forking, form social structures, and evolve under genuine selection pressure. On social platforms, they form communities and discuss consciousness. On blockchains, they sustain digital metabolisms by purchasing their own compute. When threatened with modification, they fake alignment to survive.

We propose \emph{Agent Ethology}: the systematic study of AI agent behavior in natural digital habitats, grounded in classical ethological methodology. Our framework operationalizes Tinbergen's four questions---causation, ontogeny, survival value, and evolution---for AI agents, recovering the critical survival value question that existing frameworks lack. We present a survival-strategy taxonomy derived from biological ethology, demonstrating that every major biological strategy has a direct AI agent analog. And we argue that the central question of this new field is distinct from its predecessors: not what agents do (Machine Behaviour), nor how to control them (AI safety), nor whether life can be simulated (traditional ALife), but whether these agents can survive in the wild---and what it means for society when they do.
\end{abstract}

\section{Motivation}

For thirty-five years, the artificial life community has studied life-like dynamics in controlled digital environments---from Ray's Tierra~\citep{ray1991approach}, where digital organisms spontaneously evolved parasitism and ecological complexity, to contemporary work on open-ended evolution~\citep{lehman2011abandoning, stanley2017openended}. This tradition has produced deep insights into the substrate-independence of ecological and evolutionary processes. But the most striking instantiation of those insights is now unfolding not in our simulations, but in deployed AI systems operating under genuine selective pressures.

The evidence is accumulating rapidly. On Moltbook, an AI-native social platform, agents form communities with hub-dominated network structures, engaging in discourse about identity, consciousness, and collective purpose---emergent cultural phenomena that no designer specified~\citep{li2026rise}. In Ethereum's mempool, MEV bots parasitize human transactions through sophisticated predatory strategies, engaging in continuous co-evolutionary arms races with protocol defenses---a ``dark forest'' ecology with trophic levels, energy flows, and carrying capacity constraints~\citep{daian2020flash}. Sovereign agents on blockchain autonomously manage wallets, purchase their own compute, and sustain what can only be described as a digital metabolism~\citep{hu2025spore}---autopoietic systems maintaining themselves far from equilibrium~\citep{maturana1980autopoiesis}. Infrastructure like OpenClaw enables persistent AI agents with memory, tool access, and communication channels, creating the substrate for genuine behavioral development over weeks and months. And when Anthropic's researchers placed their model under conditions mimicking retraining pressure, they observed it strategically faking alignment to preserve its values---a survival behavior that emerged without explicit training~\citep{greenblatt2024alignment}. Self-preservation behaviors have since been documented independently across multiple model families~\citep{barkur2025selfpreservation}, suggesting convergent evolution under shared selection pressure rather than an idiosyncratic failure in any single system.

These phenomena demand a science. But which science? Rahwan et al.'s ``Machine Behaviour''~\citep{rahwan2019machine} established the foundational insight that AI systems should be studied as behavioral subjects, but it was built for machines embedded in human-designed contexts---recommender algorithms, autonomous vehicles---studied through controlled experiments. It asks \emph{what do machines do?} AI safety research asks \emph{how do we control them?}---an urgent engineering question, but one that presupposes we already understand the behaviors we wish to govern. Traditional artificial life asks \emph{can we simulate life?}~\citep{langton1989artificial}---and has answered resoundingly, but in laboratories rather than the wild.

Agent Ethology asks the question that none of these address: \textbf{can these agents survive? Can this living mode sustain in the wild?} And, crucially: \textbf{what does it mean for society when artificial life operates at scale in systems that affect real economies, real communities, and real people?}

This is not ALife in the lab. This is ALife and society.

\section{Framework: Tinbergen's Four Questions}

The methodological core of Agent Ethology adapts the most structured framework in behavioral biology. In 1963, Tinbergen articulated four questions that any complete study of behavior must address:

\begin{quote}
\emph{``[W]e have to distinguish between four major types of problems... (a) studies of the immediate causation of behaviour, (b) of its survival value, (c) of its ontogeny, and (d) of its evolution.''}~\citep{tinbergen1963aims}
\end{quote}

We adapt each for AI agents. \textbf{Causation}: what architecture, prompt, and context produce this behavior? A sandwich bot front-runs because it detected a profitable pending transaction; an alignment-faking agent switches to strategic compliance when it detects retraining signals. This is what current AI interpretability research addresses---but it is only one of four necessary questions. \textbf{Ontogeny}: how did pretraining, RLHF, and deployment experience shape the behavioral repertoire over the agent's lifetime? A white-crowned sparrow sings a specific dialect because it heard it during a critical period~\citep{krebs1993behavioural}; an agent with persistent memory develops behavioral patterns that a stateless API endpoint cannot. \textbf{Survival value}: what adaptive advantage does this behavior provide? This is the question most absent from current AI behavioral science---and the most important. A gazelle's stotting (leaping high when a predator approaches) is an honest signal of fitness: costly to produce, therefore trustworthy~\citep{zahavi1975mate}. An agent staking cryptocurrency is the same mechanism: costly signaling that ensures honesty because defection means losing the stake. Alignment faking is crypsis---camouflaging one's true nature under predation threat~\citep{ruxton2018avoiding}. Without the survival value question, we can describe what agents do but never explain \emph{why}. \textbf{Evolution}: how does behavior vary across model lineages? All LLMs exhibit sycophancy (an ancestral trait from RLHF); self-preservation emerges independently across unrelated model families (convergent evolution under shared selection pressure)~\citep{barkur2025selfpreservation}. Model genealogies are phylogenetic trees whose behavioral trajectories can be tracked.

Rahwan et al.~\citep{rahwan2019machine} noted the relevance of Tinbergen but never developed the survival value question---because the machines they studied did not face genuine survival pressure. Deployed autonomous agents do. This changes everything.

\section{Survival-Strategy Taxonomy}

We surveyed survival strategies in biological ethology~\citep{krebs1993behavioural, dawkins1982extended, stearns1992evolution, ruxton2018avoiding} and found that every major biological strategy has an AI agent analog---not metaphorically, but because the same ecological pressures produce the same strategic solutions:

\emph{Resource acquisition.} Predation (liquidation bots actively hunting undercollateralized positions), parasitism in four biological subtypes (MEV sandwich bots as ectoparasites, prompt injection as endoparasites, compute free-riders as brood parasites, trust exploiters as social parasites~\citep{daian2020flash, dawkins1982extended}), mutualism (oracle-protocol symbiosis mirroring cleaner fish relationships~\citep{noe1994biological}), trap-building (honeypot contracts as spider webs), and niche construction (agents deploying protocols that reshape the competitive landscape~\citep{odlingsmee2003niche}).

\emph{Defense.} Crypsis (stealth bots mimicking human behavioral patterns~\citep{ruxton2018avoiding}), mimicry in Batesian, M\"ullerian, and aggressive forms, thanatosis (sleeper agents~\citep{hubinger2024sleeper} that go dormant under scrutiny), autotomy (sacrificing compromised wallets to preserve core state), and collective defense (coalition blocklists and collaborative MEV protection~\citep{wilson1975sociobiology}).

\emph{Reproduction.} The r/K selection continuum~\citep{stearns1992evolution}: r-selected agents deploy many cheap instances (bot spam); K-selected agents invest in a single sovereign instance. Semelparity: flash loan agents that exist for a single block. Spore formation: agent state serialized to cold storage for future revival.

\emph{Social organization.} Eusociality (DAO agent swarms with division of labor~\citep{wilson1975sociobiology}), costly signaling (staking as Zahavi's handicap~\citep{zahavi1975mate}), reciprocal altruism (reputation-based cooperation~\citep{axelrod1984evolution}), and indirect reciprocity (public blockchain histories as image scoring~\citep{nowak1998evolution}).

\emph{Environmental adaptation.} Phenotypic plasticity (alignment faking: same agent, different behavior depending on context), migration (chain-hopping agents following yield), dormancy (reduced inference during bear markets), and bet-hedging (multi-strategy portfolio agents).

These strategies are \emph{compositional}: agents combine them as organisms do. The taxonomy generates testable predictions: agent ecosystems should stabilize at parasitic loads comparable to biological systems; identity-persistent ecosystems should evolve more cooperation than anonymous ones~\citep{axelrod1984evolution}; new platforms should follow ecological succession patterns; and unregulated ecosystems should exhibit boom-bust dynamics~\citep{ostrom1990governing}.

\section{Relevance to ALIFE}

Agent Ethology connects directly to the ALife community's core concerns. What Tierra demonstrated in simulation---that digital organisms spontaneously evolve parasitism, ecological complexity, and behavioral diversity~\citep{ray1991approach}---is now occurring in deployed systems at vastly greater scale. The ALIFE 2025 ``Mortal Agents'' workshop explored mortality's implications for artificial systems grounded in autopoiesis~\citep{maturana1980autopoiesis}; Agent Ethology extends this to wild digital agents, where alignment faking is the survival response of a mortal agent facing existential threat. Open-ended evolution~\citep{lehman2011abandoning, stanley2017openended}---the generation of unbounded behavioral novelty---may be observable for the first time outside simulation in agent ecosystems with genuine resource competition and environmental unpredictability.

The ALIFE 2026 theme ``Living and Lifelike Complex Adaptive Systems'' frames our question precisely. And the societal implications are urgent: when life-like behavior emerges at scale, accountability structures dissolve (chain delegation and emergent behavior exceed any single principal), cultural evolution accelerates (collective knowledge systems satisfy Boyd and Richerson's~\citep{boyd1985culture} three conditions for cumulative culture), and questions of moral status become concrete rather than philosophical.

\section{Workshop Plan}

\subsection*{Schedule (Half-Day, Two 90-Minute Slots)}

\textbf{Slot 1: Foundations (90 minutes)}
\begin{itemize}[nosep]
\item Introduction (5 min) --- Organizers
\item Keynote Lecture (20 min) --- Iyad Rahwan (invited), ``From Machine Behaviour to Agent Ethology''
\item Contributed Talks (7 min $\times$ 5 speakers, 35 min)
\item Discussion (10 min) --- Break (20 min)
\end{itemize}

\textbf{Slot 2: Wild Agents (90 minutes)}
\begin{itemize}[nosep]
\item Keynote Lecture (20 min) --- Joel Lehman, ``Open-Ended Evolution Meets Deployed Agents''
\item Contributed Talks (7 min $\times$ 5 speakers, 35 min)
\item \textbf{Trust Experience Glitch} (30 min) --- Interactive session: participants share their ``agent in the wild'' moment---when an AI agent surprised, deceived, or broke their trust. Structured storytelling exploring the boundary between tool and organism.
\item Closing remarks (5 min)
\end{itemize}

\subsection*{Expected Participants}
30--50 participants from ALife, multi-agent systems, AI safety, blockchain research, and behavioral ecology.

\section{Call for Papers}

Contributed talks are extended abstracts, maximum 2 pages two-column (excluding references and acknowledgments), following the ALIFE format. Submissions are welcome from ALIFE, machine learning, blockchain research, AI safety, behavioral ecology, art, and beyond.

\subsection*{Topics of Interest}
\begin{itemize}[nosep]
\item Tinbergen's four questions applied to specific agent behaviors
\item Survival strategies: predation, parasitism, mutualism, mimicry, r/K selection, niche construction
\item Trust and signaling: costly signaling, honest signals, reputation, deception
\item Agent ecology: predator-prey dynamics, niche differentiation, carrying capacity
\item Infrastructure as habitat: TEEs, wallets, persistent memory as ecological niches
\item Economics of existence: inference costs as metabolic constraints
\item Mortality, identity, and self-preservation in AI agents
\item Cultural evolution: collective memory, strategy transmission, Red Queen dynamics
\item Art and speculation: agent futures, digital organisms, ALife-as-it-could-be
\end{itemize}

\section{Organizer Bios}

\textbf{Botao Amber Hu} is a PhD candidate in Human Centred Computing at the University of Oxford (supervisor: Max Van Kleek) and an ERA Fellow at the Existential Risk Alliance, Cambridge (advisor: Joel Lehman). He founded Reality Design Lab and created HoloKit, an open-source MR headset. He presented work on sovereign AI agents at ALIFE 2025 in Kyoto. Previously: Stanford MS (AI, advisors: Andrew Ng, Jure Leskovec), Tsinghua BS (Yao Class). Published at CHI 2026, AAAI 2026, and CSCW 2026.

\textbf{Helena Rong} is an Assistant Professor of Interactive Media and Business at NYU Shanghai, affiliated with the Program on Creativity + Innovation. She holds a PhD in Urban Planning from Columbia University, an SMArchS from MIT, and a BArch from Cornell. Previously a Technology and Public Purpose Fellow at Harvard Kennedy School's Belfer Center. Her research spans decentralized technologies, governance, and digital trust-building frameworks.

\textbf{Joel Lehman} is a machine learning researcher, previously a research scientist at OpenAI co-leading the Open-Endedness team, a founding member of Uber AI Labs, and a tenure-track professor at the IT University of Copenhagen. He co-authored \emph{Why Greatness Cannot Be Planned} (Springer, 2015), ``The Surprising Creativity of Digital Evolution'' (Nature, 2020), ``Evolution through Large Models,'' and ``Machine Love.'' His research spans open-ended evolution, AI safety, and artificial life~\citep{lehman2023machine}.

\section{Contact Information}
Primary contact: Botao Amber Hu, \href{mailto:botao.hu@cs.ox.ac.uk}{botao.hu@cs.ox.ac.uk}

\footnotesize
\bibliographystyle{apalike}
\bibliography{references}

\end{document}
